\section{PnO 7 Task Benchmark: Further Details}
\label{app:bench}
\subsection{Dataset descriptions}
\textbf{Knapsack} using both synthetic and real energy data, \textbf{Energy Scheduling} using the same real-world dataset, \textbf{Budget allocation} based on an internet advertising dataset, \textbf{TopK selection} on synthetic data,  \textbf{Bipartite Matching} on graphs of citation data, and \textbf{Portfolio optimization} on stock data. 

\subsection{Hyperparameter details}

For all problems, we began with 9 training runs: a grid search of noise-level $\sigma \in [0.1, 0.5, 1]$ and learning rate $[0.01, 0.005, 0.001]$. All problems used $M=25$ Monte Carlo samples of noise, except Energy Scheduling which used only 1 due to its long time to train.

\subsubsection{TopK additional hyperparameter search} 
Following the advice of \citet{cordonnierDifferentiablePatchSelection2021}, we added a linear decay schedule to $\sigma$. We tried the following $\sigma$ start and end values $0.1→0.01, 0.1→0.001, 0.05→0.005, 0.05→0.001, 0.5→0.01$, with warm-up periods of no decay lasting $0, 10,$ and 30 epochs. We used the original three learning rates $[0.01, 0.005, 0.001]$, and the presence and absence of a $\tanh$ activation on the predicted costs. The best model used $\sigma 0.05→0.001$ with a warmup period of 10 epochs, no activation, and a learning rate of 0.01.   

\subsubsection{Synthetic Knapsack additional hyperparameter search}
On this task, an additional $\sigma \in [0.05, 0.2, 0.3, 1]$ and learning rate ($[0.001, 0.5]$) search yielded no improvement. We attempted adding a $\tanh$ activation to bound the predicted costs, and experimented with a larger range of Monte Carlo samples $M\in[1, 5, 25, 50]$.  The best model used $\sigma=0.5$, learning rate = 0.01, a $\tanh$ activation, and $M=50$ Monte Carlo Samples.

\subsubsection{Energy Scheduling Additional Hyperparameter Search} The model used in the Energy Scheduling task has no bounds on predicted costs, so in the second wave of hyperparameter search we tried with and without a sigmoid to constrain predicted costs in the range $[0, 1]$. We additionally attempted $\sigma$ decay from the initial values to $0.01$. The best model used $\sigma=0.5$, a learning rate of $0.01$, and linear decay on $\sigma$ from $0.5 \xrightarrow\ {0.01}$ over 100 epochs.

\subsubsection{Portfolio Additional Hyperparameter search}

For Portfolio, we actually \textit{removed} hyperparameters. This task has informative gradients, so there is no need for perturbation. We simply skipped the perturbation and used the best learning rate from the original grid search.