We replicate the 7 primary tasks from \citet{gengBenchmarkingPtOPnO2024}'s end-to-end prediction and optimization (PnO) benchmark, using their implementation (when possible) and dataset settings. We omit their novel advertising task, as its online nature precludes many common decision-aware methods. See App.~\ref{app:bench} for task specific descriptions.

The seven tasks span a range of decision structures, formally specified in Tab.~\ref{tab:decision-problems}. There are two synthetic tasks: a knapsack with a polynomial data-generating process and a top-$K$ selection task with a cubic process. There is an additional knapsack on real-world energy data, along with an energy scheduling integer program, both built on the SEMO energy market. The bipartite matching task uses edge weights derived from the Cora citation network. Budget allocation, on Yahoo! click-through data, is the only task with a nonlinear (submodular) objective. Portfolio allocation, predicting daily asset returns from S\&P 500 data, is the only task with continuous decisions and an inherently differentiable objective.

We organize the tasks into three regimes based on whether or not the task is well specified (the model can closely represent the data generating process) and the amount of training data available. This aligns with the known empirical and analytical result that in a well-specified setting, the two-stage approach is more sample-efficient than decision focused methods \citep{elmachtoubDissectingImpactModel2025, huFastRatesContextual2022}.

We add a Differentiable Perturbed Optimizer method \citep{berthetLearningDifferentiablePerturbed2020}, which was referred to in \citet{gengBenchmarkingPtOPnO2024} but not actually benchmarked. While the benchmark authors provide some code for DPO, its implementation was incomplete so we use instead code from \citet{TueroPytorchDPO}.

The original benchmark paper performed only a minor hyperparameter search, varying learning rate alone. We expand this by also testing batch size and method-specific hyperparameters. We hypothesize that this wider tuning will affect which methods appear most successful, including the two-stage MSE baseline: batch size affects every method, not just decision-aware ones. We conduct two stages of hyperparameter search. In the first we search an entire grid of:
\begin{enumerate}
    \item Adam Learning Rate: $[0.01, 0.005, 0.001]$. The original paper explored $[0.1, 0.05]$ as well, but these were omitted to save time as many models seemed to prefer $0.005$ over $0.01$.
    \item Batch size. The original paper used full-batch gradient descent for most methods; for a few methods (SPO+, NCE, and the learn-to-rank methods) \citeauthor{gengBenchmarkingPtOPnO2024}'s benchmark defaulted to a batch size of 1. Here we explore a minibatch size of 32 (4 for bipartite matching with only 20 training examples) in addition to the default full-batch or single-instance behavior.
\end{enumerate}

This initial grid covers 3 learning rates and 2 batch sizes for each of 7 tasks and 12 methods. We select the best learning rate and batch size independently for each Task/Method pair based on validation decision regret. Then, we conduct a second hyperparameter sweep of any method-specific hyper parameters. Although this may unfairly advantage any method with additional hyperparameters as it will receive more trials, in this work we are not necessarily interested in determining the optimal method, but instead with assessing what is possible on these benchmarking tasks.

Tab.~\ref{tab:decision-problems} formally states the decision problem solved in each experiment, and Tab.~\ref{tab:methods-hyperparams} lists the per-method hyperparameter grid that Phase~2 sweeps over. \citet{gengBenchmarkingPtOPnO2024} did not sweep over these method-specific hyperparameters; for each grid, the middle value matches their reported default, and we add two values above and two below to assess sensitivity.
\input{content/chapters/dpo/new_tables_figures/decision_problems.tex}
\input{content/chapters/dpo/new_tables_figures/methods_hyperparams.tex}

\textbf{On the implicit decision-awareness of MSE.} We follow the benchmark convention of calling MSE \emph{decision-unaware}, but emphasize that every cell of our search grid is selected by minimum validation \emph{decision regret}. Grid search over a finite hyperparameter set against a decision-quality criterion is itself a (coarse) form of decision-aware learning, in the sense that information about the downstream cost function informs model selection. Using the criterion one actually cares about for model and hyperparameter selection is common practice. The MSE result reported here is therefore best read as a particularly cheap and widely-applicable decision-aware baseline rather than a strict decision-blind reference.

\textbf{Hyperparameter tuning helps decision-aware methods, but not enough to overturn the well-specified result.} Tab.~\ref{tab:tuning-benefit-summary} summarizes how much each method's test regret improves on the bench tasks once its method-specific hyperparameter is tuned, rather than fixed at \citeauthor{gengBenchmarkingPtOPnO2024}'s reported default value; the full per-task breakdown is in Tab.~\ref{tab:tuning-benefit} of App.~\ref{app:bench}. DFL, lsLTR, and Blackbox each gain $>10\%$ on average from tuning, yet on the well-specified tasks below they still rarely close the gap to MSE.
\input{content/chapters/dpo/new_tables_figures/tuning_benefit_summary.tex}

\subsection{Well-Specified Regime: MSE Wins or Ties}
We report the results of this exercise for the Knapsack (real and synthetic), Scheduling, Top-K, and Bipartite Matching tasks in Figure \ref{fig:bench_bump}. We group these 5 tasks together post-hoc, based on the empirical finding that MSE is superior to nearly all methods, save for listwise-LTR, with which it has similar performance.

We hypothesize the surprisingly strong performance of MSE is due to these problems being well-specified. The MLP model with a single 32-unit hidden layer used by the benchmark has sufficient capacity to represent the relatively simple data generating processes in these tasks. Although the energy data is real-world, it consists of hourly energy prices, which follow a predictable pattern that is easily learned by a small neural network with little distribution shift between train and test. Indeed in the Top-K (Cubic) task, we see that some methods are capable of virtually 0 prediction error on heldout data. The knapsack and scheduling tasks show a similar scenario: relative regret is between $1\%-5\%$.

The bipartite matching task is an exception. We see that LTR methods are generally capable of approximately $90\%$ relative regret, while MSE is near $92\%$. Given the large gap between the predicted decisions and optimal, this task may be too difficult for any extant method with only 20 training examples. The sparse bag-of-words features available are also not informative enough to learn the citation patterns.

\subsection{Misspecified Regime: Budget Allocation}
\label{sec:bench-budget}
Results on the Budget Allocation task are shown in Figure~\ref{fig:budget_bump}. The task is to allocate a fixed budget across websites to maximize the submodular probability that a user clicks on at least one advertisement, using real-world click-through data. Here we see an opposite story to the well-specified regime: MSE is worse than all but one decision-aware method. MSE only achieves $35\%$ relative regret, while the best decision-aware methods have less than $5\%$ on heldout data. We argue this task is much more informative for decision-aware model selection than the rest, because it is the only task where decision-aware models have clear advantages over MSE. We hypothesize that this is due to two factors. The data-generating process for click-through rates is complex and real-world. The cost function is also nonlinear (submodular), qualitatively different from the linear costs used in the other benchmark tasks.

\subsection{Continuous, Gradient-Informative Task: Portfolio}
\label{sec:bench-portfolio}
Results on the Portfolio task are presented in Figure \ref{fig:portfolio_bump}. These are presented separately because we believe this task should not be included in benchmarks of decision-aware methods without sufficiently noting that this is not a combinatorial optimization task. This task is to optimize the allocation of funds in an investment portfolio, and the allocation fractions are treated as continuous decision variables. Accordingly, the objective is not piecewise constant with respect to model parameters and has informative gradients.

When originally used as a decision-aware benchmarking task in \citet{shahDecisionFocusedLearningDifferentiable2022}, it was proposed as a "stress-test," in that it was a task where decision-aware methods were known to preform poorly. Here \citeauthor{shahDecisionFocusedLearningDifferentiable2022} show how their novel decision-aware method (LODL) performs at the same level as MSE. Furthermore, these authors show that the best method in this case is simply optimizing the objective directly. Here, this is called Decision Focused Learning (DFL), in the generic sense of \citeauthor{shahDecisionFocusedLearningDifferentiable2022}, distinct from the DFL by Wilder et al.\ method discussed above. While \citeauthor{gengBenchmarkingPtOPnO2024} implement a DFL method in the benchmark, it \textit{always} uses a surrogate loss, even though this is not necessary on the Portfolio task. We remedy this, and highlight the strong performance of DFL on this task in Figure \ref{fig:portfolio_bump}.

