We hope this work enables future researchers designing decision-aware learning techniques and practitioners implementing them to better understand and evaluate the strengths of different models.
In particular, we have provided evidence that:
\begin{itemize}
    \item Many tasks commonly used to benchmark decision-aware methods are too simple or too well-specified to benefit from decision aware learning. Two-stage, decision-blind approaches should be attempted and given sufficient hyperparameter tuning budget (noting that selecting hyperparameters by validation decision regret is itself a coarse form of decision-aware learning, and that even MSE-trained MLPs are sensitive to batch size because the network is non-convex in $\theta$). As for particular tasks: the Budget Allocation task appears to be a good candidate for comparing decision-aware learning methods, but the best methods currently achieve low ($<5\%$) regret, suggesting that future research will likely struggle to improve meaningfully on the current results. The Bipartite Matching task could possibly be a revealing task, as the best methods in this task still achieve roughly $90\%$ regret. Our preliminary experiments suggest performance well above this floor is attainable with modest changes to the model class. Finally, the Portfolio task is not combinatorial optimization. We caution against its use without sufficiently noting this and separating the interpretation of results on this task from other decision aware tasks.
    \item Through analysis of the synthetic data benchmark of \citeauthor{huang2024decisionfocusedPG}, we have separated the quality of a decision-aware loss function from the quality of gradient-based learning under that loss, captured by two complementary dimensions: ``What solution does this method prefer?" and ``What solution is this method capable of learning?''. We have also demonstrated how varying key hyperparameters can change the answer to both of these questions.
    \item Finally, we revisit the real-world top-K site-selection comparison of Chapter~\ref{chap:daml}. The apparent advantage of direct BPR optimization (``DPO'') over surrogate methods on these tasks reflects training-objective alignment with the evaluation metric: BPR is a normalized per-instance objective, while off-the-shelf PG and SPO+ surrogates target the unnormalized linear decision loss. Once aligned, the surrogate landscapes resemble BPR's, suggesting the methods do not differ intrinsically on this kind of task but in their default training objective.
\end{itemize}

\textbf{Limitations.} The largest limitation of this work is the lack of uncertainty quantification on our reported regrets and BPR values. A larger run with more random seeds and trials per task/method/hyperparameter cell would let us determine which differences between methods are statistically significant and which are within trial-to-trial noise. Beyond this, our conclusions are constrained by the model classes and data scales studied. The PnO benchmark uses a single 32-unit MLP across tasks; richer architectures or feature sets could shift which tasks appear practically well-specified. The synthetic study uses a 2-parameter linear model where any approximate match of the decision threshold yields near-oracle decisions, so we cannot say from this alone how PG's tolerance to landscape bias scales to higher-dimensional models. Finally, our top-K analysis examines a single model class (Negative Binomial regression with random effects) on two real-world domains, so the extent to which surrogate-method differences are explained by training-objective alignment in other top-K settings remains an open empirical question.