In these problems, a decision $z\in\mathcal{Z}$ must be chosen from a
discrete feasible set to minimize a cost that depends on outcomes $Y$ that are
unknown at decision time. We model $Y$ as a random variable produced from a
conditional distribution given observed features $X=x$. 
For example, in a later top-K site selection tasks each $z$ represents an indicator vector that picks $K$ sites for intervention out of many, while vector $y$ has an observed score for each site, with smaller values indicating desirable sites given $x$. \
This notation is reversed from previous chapters, where a higher score meant a better opportunity to intervene on more overdoses or observe more wildlife. \
Here we want to minimize a total cost and act on the places with the lowest scores. \
To put it in terms of our grounding public health example: the human cost of lives lost is minimized when we intervene on the places with the most overdose deaths. To convert deaths to a suitable score, we could simply negate the count. \
%hospitalized patient monitoring application, $z$ may represent a subset of patients selected to receive additional clinical attention from among all feasible subsets $\mathcal{Z}$, while $y$ indicates survival or death in next 24 hours for all patients, given patient-specific attributes $x$. 
We write that  the decision to do action $z$ when the outcome is $y$ incurs a scalar cost $c(z,y)$. 

The decision-making problem is, given an $x$, to solve for the value(s) of $z$ with lowest expected cost,\begin{align}
%% OLD VERSION:
% z^\star(x)\;\in\;\arg\min_{z\in\mathcal{Z}} \mathbb{E} \! \left[c(z,Y)\mid X=x\right].
% \\ \notag 
%% MCH PROPOSED:
z^\star(x,\theta)\;:=\;\arg\min_{z\in\mathcal{Z}} \mathbb{E}_{p_{\theta}(Y | X=x)} \!\left[c(z,Y) \right].
\label{eq:decisions}
\end{align}
Here, the parameter $\theta \in \mathbb{R}^P$ defines a conditional distribution over $Y$ given $X{=}x$.
At decision-making time (``test time''), only features $x$ and parameter $\theta$ are available; the realization of $Y$ is
unknown until later.

\textbf{Linear cost assumption.} A common assumption \citep{berthetLearningDifferentiablePerturbed2020, elmachtoubSmartPredictThen2022, huang2024decisionfocusedPG} is that cost is a linear inner product $c(z, y) = 
\langle z, y\rangle$. This is to say that the optimization problem can be written as a linear program (LP).
We note that this is not a restrictive assumption: for any score function $s$ which maps a decision vector $z \in \mathcal{Z}$ to a real-valued scalar, we can write $\max_{z \in \mathcal{Z}} s(z)$ as a linear program for some embedding of $\mathcal{Z}$ \citep{berthetLearningDifferentiablePerturbed2020}, though this embedding may be nonlinear and grow with $|\mathcal{Z}|$ (and need not be known explicitly to apply these methods, given a blackbox optimization oracle). As a worst case, a vacuous embedding places each feasible decision on its own dimension, recovering an LP over the simplex of $|\mathcal{Z}|$ vertices.

For most of this chapter we will assume a linear cost objective; the notable exception is the budget allocation task in Section~\ref{sec:bench-budget}, which has a submodular (nonlinear) cost.
In the linear-cost case, $z$ and $y$ are vectors of size $D$, where $D$ is a problem-specific dimension (e.g., the number of edges in bipartite matching or the number of sites in top-$K$ selection). Linearity of the cost in $y$ allows the expectation to collapse: $\mathbb{E}[\langle z, Y\rangle\mid X{=}x] = \langle z, \hat{y}_{\theta}(x)\rangle$, where $\hat{y}_{\theta}(x)$ denotes the mean of $Y$ under $p_{\theta}(Y | X=x)$. Decisions then depend only on the conditional mean, and the inner minimization reduces to $z^*(x,\theta) = \arg\!\min_{z \in \mathcal{Z}} \langle z, \hat{y}_{\theta}(x) \rangle$, a linear program. Linearity in $y$ alone is enough for this expectation step. We note that the portfolio task in Section~\ref{sec:bench-portfolio} has a cost $\langle z, y\rangle - z^\top A z$ that is linear in $y$ but quadratic in $z$, so the inner minimization is a quadratic program rather than an LP. However, the surrogate methods discussed below derive their gradient signals under the linear-cost assumption. On portfolio, their training signals capture only the linear-in-$y$ part of the decision cost (the realized return given $z^*(\hat y)$); the quadratic risk term $\alpha z^*(\hat y)^\top \Sigma z^*(\hat y)$ contributes to the reported test regret but not to their gradient. We return to this caveat in Section~\ref{sec:bench-portfolio}.
In all of these cases (linear, quadratic, and submodular), we see the key challenge of decision-aware learning: the decision problem is piecewise constant as a function of $\hat{y}_{\theta}$. Its flatness yields all-zero gradients which break gradient-based learning of $\theta$.

\subsection{Training: Decision-unaware vs Decision-aware}

The training problem is to estimate $\theta$ given observed data $\{(x_i,y_i)\}_{i=1}^n$ assumed drawn i.i.d.\ from the joint distribution of $(X,Y)$.

\textbf{Decision-unaware training.}
One natural choice is to learn $\theta$ to maximize the likelihood $p_{\theta}(Y|X)$ given the $N$ observed data pairs. Under a Gaussian likelihood, this corresponds to minimizing the mean squared error (MSE):
\begin{align}
    \theta^* \gets \arg\!\min_{\theta} \frac{1}{N} \sum_{n=1}^N (y_n - \hat{y}_{\theta}(x_n) )^2.
\end{align}
Here, neither decision variable $z$ nor the cost function $c$ has any direct role in determining $\theta$. This approach is called MSE by \citeauthor{gengBenchmarkingPtOPnO2024}'s benchmark.

This estimation strategy can be used in a simple two-stage approach for decision-making. First, estimate $\theta$ by minimizing MSE. Once $\theta$ has been estimated, in the second stage we make instance-specific decisions via Eq.~\eqref{eq:decisions}. We use ``two-stage'' to denote the decoupled training pipeline, where $\theta$ is fit without reference to the cost $c$. Decision-aware methods, by contrast, fold $c$ into training even though they too solve Eq.~\eqref{eq:decisions} at test time.

When the probabilistic model is \emph{well-specified}, meaning the chosen parametric model class contains the true data-generating process, this two-stage approach is well-founded. We cannot in general directly verify this condition on real-world tasks; in Sec.~\ref{sec:bench_replication} we instead infer well-specification operationally from the regret gap between two-stage MSE and decision-aware methods, following \citet{elmachtoubDissectingImpactModel2025}. In fact, it may have two distinct advantages over decision-aware methods. First, it is a simple way to bypass the lack of informative gradients. Second, it can achieve fast convergence on regret when small estimation errors do not change the optimal decision \citep{huFastRatesContextual2022}, and can have smaller high-regret probabilities than decision-aware training when the predictive model is well-specified or nearly well-specified \citep{elmachtoubDissectingImpactModel2025}. However, for interesting real-world decision problems, our parametric model $\hat{y}_{\theta}(x)$ will always be misspecified, especially if we select a simpler model class to address interpretability or data scarcity. In these more realistic cases, training with a decision-aware objective can lead to better downstream decisions.

\textbf{Decision-aware training.}
End-to-end decision-aware methods seek an estimated $\theta$ that is deliberately steered to deliver good decisions via Eq.~\eqref{eq:decisions}. Methods that pursue this kind of objective are interchangeably called decision-aware learning, decision-focused learning, or predict-and-optimize (PnO) in the literature. This approach to training directly seeks to minimize the decision cost
\begin{align}
    \theta^* \gets \arg\!\min_{\theta} ~~ \frac{1}{N} \sum_{i=1}^n c(z^*(x_i,\theta), y_i).
\end{align}
This generic parameter estimation framework faces the technical hurdle that its cost $c$ is piecewise constant as a function of $\theta$. Gradients of $c$ with respect to $\theta$ vanish almost everywhere, so stochastic gradient descent cannot directly be used to solve for $\theta^*$. Below, we survey a variety of approaches to overcoming this hurdle.

We will often report results in terms of \emph{regret}: the gap between the cost incurred by the model's decision and the cost of the oracle decision that knows the true outcome,
\begin{align}
\text{regret}(x, \theta, y) := c(z^*(x, \theta), y) - \min_{z' \in \mathcal{Z}} c(z', y).
\end{align}
The oracle decision $z'$ is obtained by calling the optimization solver with the true $y$. For a given model class, it may not be possible to predict any $\hat{y}$ that recovers $z'$. To compare methods on a dataset of $n$ observations, we report \emph{relative regret}, the average regret normalized by the average oracle cost:
\begin{align}
\textstyle
\text{rel-regret}(\{x_i,y_i\}_{i=1}^n, \theta) := \frac{\sum_{i=1}^n \text{regret}(x_i, \theta, y_i)}{\sum_{i=1}^n |\min_{z' \in \mathcal{Z}} c(z', y_i)|}.
\end{align}

\subsection{A taxonomy of Predict-And-Optimize methods}
We will use the same taxonomy of methods used by \citeauthor{gengBenchmarkingPtOPnO2024}, which we will briefly describe here. Some taxonomy category names have been changed to better clarify their qualities.

\textbf{Surrogate Gradients}
One family of PnO methods preserves the discrete optimization problem, but somehow augments the problem such that the backward pass produces informative gradients. 
This family is referred to as \textbf{Discrete} by \citeauthor{gengBenchmarkingPtOPnO2024}, but we think it is more informative to name this group of methods after how they deal with the challenges of decision-aware learning.
The benchmark of \citet{gengBenchmarkingPtOPnO2024} explores \textbf{Blackbox} \citep{pogancicDifferentiationBlackboxCombinatorial2019}, which works by linearly interpolating the piecewise-constant gradient with hyperparameter $\lambda$ to control how informative the gradient is, and \textbf{Identity} \citep{sahooBackpropagationCombinatorialAlgorithms2022}, which replaces the backwards pass through the combinatorial solver with the identity. It also explores \textbf{DFL by Wilder et al.} \citep{wilderMeldingDatadecisionsPipeline2019}, which does not differentiate through the combinatorial solver, only the calculation of the cost after the solver returns a decision, regularized by mean-squared error with weight $\alpha$. We use the qualified name ``DFL by Wilder et al." to distinguish this specific method from the broader meaning of ``DFL'' (decision-focused learning) used by \citet{shahDecisionFocusedLearningDifferentiable2022} and adopted later in this chapter.

\textbf{SPO+} \citep{elmachtoubSmartPredictThen2022}, the Smart Predict-then-Optimize loss, is a convex upper bound on regret. Although \citet{gengBenchmarkingPtOPnO2024} claim to use SPO-relax (a continuous relaxation of SPO+), the provided code in fact uses SPO+, and we use SPO+ as well. SPO+ differs slightly from the other Surrogate Gradient methods above: rather than substituting a new gradient for the flat one, SPO+ constructs a statistically bounded surrogate objective whose gradient is naturally non-zero. As we discuss in Section~\ref{sec:methods_added}, PG behaves similarly.

\textbf{Statistical Methods}
These techniques work by replacing the optimization problem with a comparison of sampled solutions. Here Noise Contrastive Estimation \textbf{NCE} \citep{mulambaContrastiveLossesSolution2021} and Learn To Rank \textbf{LTR} \citep{mandiDecisionFocusedLearningLens2022} both work by maintaining a ``solution cache`` of sampled solutions to the optimization problem. Then, the training signal is replaced by a loss function designed to capture the quality of the solution and not necessarily the downstream decision regret.  The listwise version of LTR introduces a temperature hyperparamter to control the smoothness of the loss function.

\textbf{Trained Surrogates} The final group of the taxonomy works by replacing the optimization problem with a flexible neural network that learns a surrogate solution. The sole example of this family is Locally Optimized Decision Losses \textbf{LODL} \citep{shahDecisionFocusedLearningDifferentiable2022}. While this method introduces a host of hyper parameters (the entire design and training of the surrogate neural network), in our later experiments we will only consider the number of training samples used to train the replacement network. All other LODL hyperparameters are set to the defaults from \citet{gengBenchmarkingPtOPnO2024}.

\subsection{Other methods considered in this chapter}
\label{sec:methods_added}
While the benchmark of \citet{gengBenchmarkingPtOPnO2024} is thorough, it is not exhaustive. In this work we will consider two additional predict-and-optimize methods: Differentiable Perturbed Optimizers (DPO) \citep{berthetLearningDifferentiablePerturbed2020} and Perturbation Gradient \citep{huang2024decisionfocusedPG}. DPO is mentioned by \citet{gengBenchmarkingPtOPnO2024}, but not included in their experiments, and PG was published contemporaneously with the benchmark so could not have been included. Placing these methods into the taxonomy of \citet{gengBenchmarkingPtOPnO2024}, DPO is a Surrogate Gradient method: the discrete solver is used but perturbed to obtain informative gradients in spite of the flatness of $c$ as a function of $\hat y_\theta$. PG could likely also be placed in the Surrogate Gradient category as it still relies on calling the discrete optimization solver. However, one could argue that PG along with SPO+ belong in their own category of designed surrogate objectives. These methods replace the decision objective with a bounded, differentiable replacement.

\textbf{DPO} The differentiable perturbed optimizer~\citep{berthetLearningDifferentiablePerturbed2020} addresses the flat-gradient problem by smoothing the decision map via random perturbations of the predictions. For $D$-dimensional noise vector
$\xi \in \mathbb{R}^D$ drawn from a standard Normal or Gumbel distribution and noise-level $\sigma>0$, the perturbed decision is defined as
\begin{align}
z_\sigma(x, \theta)
\;:=\;
%\textstyle
\mathbb{E}_{\xi}[
~
\arg\min_{z\in\mathcal{Z}} \langle z,\, \hat{y}_{\theta}(x) + \sigma \xi\rangle
~].
\end{align}
In practice, we can estimate this perturbed decision vector of size $D$ and its gradient via Monte Carlo estimates. Throughout the rest of this work, we focus on iid Gaussian noise and assume $M$ samples, each $\xi^{(m)} \sim \mathcal{N}(0, I_D)$ a $D$-dimensional vector. Define the decision obtained with each noise sample as as $\hat{z}^{(m)} {=} \arg\min_{z\in\mathcal{Z}} 
~\langle z,\, \hat{y}_{\theta}(x) + \sigma \xi^{(m)} \rangle$, then we can estimate the perturbed decision $\hat{z}$ and the Jacobian matrix $J$ needed for gradients as
\begin{align}
\hat{z}(x, \theta)
&=
\textstyle 
\frac{1}{M} \sum_{m=1}^M
\hat{z}^{(m)}
\\ \notag 
\hat{J}(x, \theta) &= 
\nabla_{\theta} \hat{y}(x,\theta) \Big[ 
\textstyle
\frac{1}{\sigma M} 
\sum_{m=1}^M \textsc{outer}(\hat{z}^{(m)}, \xi^{(m)})
\Big] 
\end{align}
This Jacobian estimator 
$\hat{J} \approx \nabla_{\theta} {z}_{\sigma}(x, \theta)$, a $P\times D$ matrix, comes directly from the chain rule and Prop. 3.1 of \citeauthor{berthetLearningDifferentiablePerturbed2020}~Noise distributions other than Gaussian require changes to the outer product term.

\textbf{DPO hyperparameters.}
DPO's most sensitive hyperparameter is noise-level $\sigma$, which controls a tradeoff between bias and gradient informativeness. As $\sigma \rightarrow 0^+$, the perturbed decision approaches the unperturbed $z^*(x,\theta)$ in expectation, so bias vanishes. However, the Monte Carlo gradient estimator becomes uninformative: in the limit we recover the piecewise-constant decision map with zero gradients almost everywhere, and at small $\sigma$ most perturbations fail to shift the $\arg\min$ at all. Larger $\sigma$ induces smoother decision maps with more informative gradients, at the cost of biasing decisions away from $z^\star$.
%Desireably, even with a single sample and small $\sigma$ the Jacobian will be non-zero.

As typical in Monte Carlo methods, DPO's estimators become more accurate as the number of samples $M$ increases: larger $M$ produce lower-variance estimates at the cost of an increased number of calls to the (often expensive) optimization solver that produces $\hat{z}^{(m)}$.

\textbf{PG method.} 
The perturbation gradient method~\citep{huang2024decisionfocusedPG} addresses the flat-gradient problem by replacing the decision loss with a smooth surrogate derived from finite differences.
Define the optimal value function for an arbitrary predicted cost $t$ as $V(t) := \min_{z \in \mathcal{Z}} \langle z, t \rangle$.
By Danskin's theorem, the decision loss $\ell(t, y) = \langle z^*(t), y \rangle$ equals the directional derivative of $V$ at $t$ in direction $y$.
The PG surrogate approximates this derivative via backward differencing with step size $h > 0$:
\begin{align}
\hat{\ell}_h(\hat{y}_\theta(x),\, y)
\;:=\;
\tfrac{1}{h}\Big(
  V(\hat{y}_\theta(x)) - V(\hat{y}_\theta(x) - h\, y)
\Big).
\end{align}
The gradient of this surrogate with respect to the predictions is the scaled difference of two optimization solves:
\begin{align}
\nabla_{\hat{y}} \hat{\ell}_h(\hat{y}_\theta(x), y)
= \tfrac{1}{h}\Big(z^*(\hat{y}_\theta(x)) - z^*(\hat{y}_\theta(x) - hy)\Big),
\end{align}
where each $z^*(\cdot)$ requires one call to the combinatorial solver. Gradients with respect to $\theta$ then follow by the chain rule. 

\textbf{PG hyperparameters.}
PG's key hyperparameter is the step size $h$, which controls a bias--smoothness tradeoff analogous to DPO's $\sigma$. As $h \to 0$, the surrogate approaches the true decision loss, but its gradient varies more sharply across nearby inputs, so the loss landscape becomes rougher and harder for gradient descent. Larger $h$ yields a smoother, easier-to-optimize loss at the cost of a more biased approximation. \citeauthor{huang2024decisionfocusedPG} recommend $h = O(n^{-1/2})$ when the feasible set $\mathcal{Z}$ is polyhedral (a separate recommendation is provided for general linear hypothesis classes which also depends on problem dimensionality \citep[Theorem 3.4]{huang2024decisionfocusedPG}).

\textbf{Non-convexity and initialization.}
The DPO and PG surrogate losses are non-convex in their predictions $\hat{y}$, and therefore non-convex in the model parameters $\theta$ for any prediction model $\hat{y}_\theta(x)$. Gradient-based training of these methods can converge to a local minimum that depends on the initial $\theta$. Common mitigations include warm-starting from MSE-trained or SPO+-trained parameters, or running multiple random initializations and selecting by validation decision regret. We adopt a single random initialization throughout this chapter; the local-minimum question is revisited empirically in Section~\ref{sec:synth} and on the real-world Top-K task.



% at added computational cost. The number of samples $M$ controls the variance of the Monte Carlo approximation; a larger number of samples provides a lower-variance decision at the cost of an increased number of calls to the (often expensive) optimization solver.

% For either the ideal expectation or the M-sample estimator, as long as $\sigma$ is large enough that some different $z$ are chosen in the inner $\arg\!\min$, the decision will not suffer from the ``flatness'' problem, thus allowing informative (non-zero) gradients to be backpropagated through the entire computational graph.




%\subsection{Gradient SNR}

%Max Welling 2020 MRI NeurIPS, 2008 weight perturbation NeurIPS.