\chapter*{Abstract}
In this thesis, I demonstrate how machine learning models can be trained and evaluated to produce better decisions when allocating limited resources under uncertainty.
I work in the predict-then-optimize setting, in which predictions from trained models are used to solve downstream discrete optimization problems, such as selecting a budget-constrained subset of locations for intervention.
While standard machine learning models are trained to minimize prediction error, this objective can be misaligned with decision quality when the model is misspecified or data is insufficient; correctly identifying the highest-priority locations matters more than minimizing average error across all locations.
Decision-aware learning methods address this misalignment by training models to optimize decision quality jointly with the prediction model, but open challenges remain: evaluation metrics often fail to capture decision-relevant performance, differentiating through discrete decision problems is difficult, and practitioners lack clear guidance on which methods to use. 
This thesis addresses these challenges through three contributions spanning evaluation, training, and benchmarking of decision-aware models.

First, I introduce Percentage of Best Possible Reach (\%BPR), an intervention-aware evaluation metric that measures how well a model's recommended resource allocation compares to the best possible allocation in hindsight.
Applied to spatiotemporal forecasting of opioid-related overdose deaths, \%BPR reveals that top-performing models by standard error metrics can miss meaningful numbers of reachable overdose events.
Second, I develop Decision-Aware Maximum Likelihood (DAML), a training method that jointly optimizes probabilistic likelihood and decision quality by using perturbed optimizers to overcome the zero-gradient barrier inherent in discrete selection problems.
DAML delivers high-quality top-K rankings while maintaining good probabilistic forecasts, and I demonstrate its effectiveness on public health and ecological monitoring tasks.
Third, I analyze a systematic benchmark of decision-focused learning methods, where I find that many common benchmarking tasks do not benefit from decision-aware learning. 
I make recommendations on tasks which are better suited to evaluating decision-aware methods, and I provide a lens with which practitioners and modelers can analyze the performance of their chose methods.
Collectively, these contributions provide researchers and practitioners with tools to evaluate, train, and select decision-aware models for resource allocation problems.