\chapter*{Acknowledgements}

This work would not be possible with the guidance and mentorship of my advisor, Dr. Michael C. Hughes. I am immensely greatful for the positive environment of his lab, the opportunity to pursue rigorous machine learning alongside meaningful applications in health, and his constant support. 

I would also like to thank the rest of my dissertation committee. To Dr. Liping Liu, I am greatful for his education in deep learning and collaborations early in my career. I will always remember the year I spent studying with Dr. Remco Chang; I am grateful for the way he allowed his students to lead him into unknown research directions. Dr. Shikhar Shrestha made it possible for me to pursue the public health research I am passionate about and gave me countless opportunities for academic and career advancement. Finally, I am thankful for Dr. Michael Huang and his gracious invitation to encounter new fields of scientific endeavor. 

This thesis would not be possible without the support of my labmates. I thank Dr. Kaitlin Gili for the joy she brought to our group, and helping me rediscover the fun in research work. I am glad to have shared a lab with the future doctors Cynthia Feeney and Ethan Harvey. Cynthia's compassion will always stick with me, I am inspired by her dedication to understanding the ways our research can help and harm vulnerable populations. Ethan is a paragon of Bayesian thought and I have benefitted immeasurably from both his knowledge, friendship, and eagerness to share his passion with others. I also thank Dr. Preetish Rath and Dr. Zhe Huang for their companionship and warm welcome at the start of my PhD career.

My academic journey has been a long one and I am indebtted to innumerable educators throughout my life. I would like to specifically acknowledge Dr. Gilad Lerman for the generous mentorship. I would not be where I am today without all of the time and resources he invested in me when I was only a highschool student. My passion for research in health stems from the guidance of Dr. Carrie Salafia and I would like to thank her for showing my how the computational work I love can make lives better.  I am also greatful to Dr. Irina Rish for the opportunities and mentorship she provided me, inspiring my love of machine learning.

Finally, I must acknowledge those who have sacrificed the most to make this work possible, my family. My PhD has come at great expense of both time and money, but my wife, Amy, has shown me a seemingly endless well of love and support throughout these years. I would not have even begun this journey without her love and encouragement, let alone finished it. Lastly, I must acknowledge the support of my mother, Lynne. From encouraging my curiousity as early as I can remember (I don't know how she figured out how a fax machine worked before Google, but she got the answer for a 4-year old me), to leaving work early to drive me to math class for years, to uprooting her life and spending her retirement supporting me by taking care of her grandson. Thank you mom.
