In this dissertation, we developed robust machine learning models aimed at improving the monitoring for hospitalized patients. This exploration is particularly crucial given the typical challenges faced in clinical settings, such as data missingness, lack of detailed labels, and high rates of false alarms. These factors often hinder the effective deployment of machine learning technologies in healthcare.

Throughout the thesis, we introduced methodologies and models designed to address these pervasive issues. Chapter 3 begins with the optimization of early warning classifiers to control false alarms through a precision constraint, enhancing their ability to mitigate alarm fatigue \citep{sendelbach2013alarm} and reduce distrust on early warning systems amongst clinicians. Following this, in chapter 4 we introduce a semi-supervised Hidden Markov Model to effectively handle time series data with missing values, proving its efficacy in realistic hospital scenarios. Further, this thesis explores the management of missingness in irregular time series, a common occurrence in EHRs, by proposing a generative model that learns the data-dependent missingness mechanisms.

The empirical evaluations presented demonstrate the significant improvements these models provide over traditional approaches, particularly in their ability to handle large-scale health records datasets effectively. 
We also leverage some of our models to not only deliver improved performance in predicting patient outcomes but also provide interpretable insights, which are crucial for clinical decision-making.
% We also leverage some of our models to not only show enhanced performance in predicting patient outcomes but also offer interpretable insights, which are crucial for clinical decision-making.

The methodologies presented in this thesis enhance the accuracy and reliability of patient monitoring systems, potentially leading to better patient monitoring through timely and precise medical interventions. These innovations pave the way for future research to explore further integrations of machine learning in healthcare, focusing on transparency, scalability, and fairness across diverse patient cohorts.

\subsection{Limitations and Future Work}
While the methodologies developed in this thesis represent significant advancements in healthcare monitoring using machine learning, they come with inherent limitations that may impact their deployment and effectiveness in clinical settings.

\textit{Interpretability to enhance adoption by clinicians}:
While we make efforts to incorporate interpretability via the PC-HMM's learned states, it remains challenging to interpret the underlying dynamics learned by the continuous time models like the CRU-FM. While extrapolative performance improves with the CRU-FM, it remains unclear as to which specific channels would benefit from the data-dependent missingness assumption. Additionally, while models adhere to pre-specified false alarm rate on the entire population via our sigmoid objective, the performance on individual subgroups remains under-explored. Further questions remain unanswered such as what specific channel at which specific time point triggers an alert? Incorporating methodologies from the XAI literature \citep{tonekaboni2020went, enguehard2023learning, lakkaraju2017interpretable} would enable adoption of these technologies, as healthcare providers may relate the model's interpretation to their clinical knowledge.

\textit{Missing at random assumption for missing data in irregular time series}:
Although the CRU-FM address missing data and irregular time series, the assumptions made (such as Missing At Random (MAR)) may not always hold true in real-world clinical environments. This discrepancy can lead to biases in predictions and affect the models' effectiveness in live clinical settings. Moreover, missingness may sometimes depend on unobserved data, termed as missing-not at random (MNAR). Future work could extend probabalistic models such as those proposed by \cite{kim2023probabilistic} and \cite{fouladgar2020novel} to handle irregular timing of measurements. 


\textit{Generalizability and fairness across diverse population sub-groups}:
To enhance the robustness and applicability of the models across varied healthcare environments, future research should prioritize cross-validation studies that span diverse medical settings. This approach will help to ensure that the models perform consistently well across different populations and clinical practices. Additionally, it is crucial to integrate fairness directly into the model development process. Techniques such as model desensitization \citep{boughorbel2021fairness} and the incorporation of fairness-specific constraints \citep{agarwal2018reductions} are essential to ensure that these models serve all patient sub-groups equitably, promoting widespread clinical adoption and trust in machine learning-driven healthcare solutions.


\textit{Optimizing Compute Utilization in Model Training}:
The methodologies developed in this research can sometimes converge to local optima, necessitating the use of robust computational infrastructure capable of parallel processing across multiple cores for training with various random initializations. This dependency could be mitigated through smarter initialization strategies that reduce the reliance on extensive hardware resources. For instance, our experiments demonstrated that initializing models using binary cross-entropy before transitioning to a sigmoid-based objective significantly enhanced performance compared to starting with random initializations. Furthermore, adopting programming frameworks like JAX \citep{jax2018github} instead of TensorFlow for models like the PC-HMM can substantially improve model training in resource-constrained environments. JAX offers a lower memory footprint and superior compilation efficiency, which are crucial in settings with limited hardware capabilities. By streamlining the computational demands through these approaches, future research can focus on creating more accessible and efficient modeling techniques that maintain high performance without requiring extensive computational infrastructure. This shift not only optimizes resource usage but also extends the feasibility of deploying advanced machine learning models to a broader range of environments.

\textit{Upgrading models to incorporate various modalities of data}:
The focus of our current models is predominantly on processing standard measurements like laboratory results and vital signs. However, there is substantial potential for future enhancements that could integrate a wider range of data modalities. Models like RAIM \cite{xu2018raim} offer promising frameworks that could be adapted to include high-frequency data streams such as electrocardiograms and pulse plethysmographs. Additionally, expanding these models to incorporate other essential Electronic Health Record (EHR) components like medications, clinical notes, and imaging data—such as X-rays and CT scans—could dramatically enrich the model's utility and accuracy.

Through addressing these limitations and exploring these avenues for future research, the field can move towards more robust, interpretable, and equitable healthcare solutions that leverage the full potential of machine learning in hospitalized patient monitoring.