The rise of big data and machine learning (ML) has the  potential to improve the quality and timeliness of medical care for hospitalized patients. Machine learning is a branch of artificial intelligence that involves training algorithms to recognize patterns and make decisions based on data. Given a time series of recent lab results and vital signs, an ML system can make personalized predictions of   short-term risk of  in-hospital mortality or onset of a life-threatening disease. 

Three key challenges stand in the way of effective ML models in these applications: 

\begin{itemize}
    \item High False Alarm Rate: Deployed models frequently generate an excessive number of false alarms, undermining their utility and trustworthiness in clinical settings.
    
    \item Missingness: At each moment an observation is made,  most labs and vitals are not updated  either due to differences in frequency of measurements or due to the caretaker's judgment of the patient’s condition. Common open-access electronic health records (EHR) datasets, such as MIMIC-III \citep{pollard2016mimic}, PhysioNet \citep{goldberger2000physiobank}, and eICU \citep{eICU} contain missingness rates of  >80$\%$ \citep{che2018recurrent}.

    \item Limited Labels: Supervisory labels for every sequence may be  limited due to the expense and time investment needed to obtain accurate disease annotations from medical staff. Several disease cohorts have missing labels in open-access electronic health records (EHR) datasets.
\end{itemize}

The existing solutions to these problems in the modern analysis of EHRs are inadequate. Existing deep models handle feature missingness by either using overly simplistic imputation strategies \citep{purushotham2018benchmarking, harutyunyan2019multitask} or by imputing deterministically without uncertainty \citep{che2018recurrent, cao2018brits}. Moreover, these models often neglect data-dependent  missingness patterns (for e.g. certain lab measurements are only collected for critically ill patients), are sensitive to the initialization of missing values \citep{cao2018brits}, or require hundreds of thousands of parameters with a lack of clinical interpretability \citep{cao2018brits, purushotham2018benchmarking, harutyunyan2019multitask}. Later, when these models are deployed, they often lead up to 90$\%$ of all alarms being false alarms \citep{fernandes2020detecting}. Another common issue is that ML models rely heavily on labels and obtaining precise annotations of disease labels from clinicians is often laborious and expensive. 

The goal of my thesis is to address these 3 barriers to deploying ML prediction systems in clinical care settings. I will introduce new probabilistic models to handle missing data, limited labels and new paradigms for training to address false alarms and unlabeled data.
 
First, in work published at AISTATS 2022 \footnote{Code URL: \href{https://github.com/tufts-ml/false-alarm-control}{github.com/tufts-ml/false-alarm-control}}, I introduced a novel training objective that can directly constrain false alarm rates in binary classifiers \citep{rath2022optimizing}. This objective defines differentiable bounds on the total counts of true and false positives that are tighter than previous work, enabling effective gradient-based learning. Clinical prediction models trained with this objective require false alarms rates to be below a tolerable value (specified by physicians), thus ensuring that clinical resources are directed at patients who are truly at-risk. When predicting in-hospital mortality on two large EHR datasets, experiments showed that our proposed objective led to absolute gains of 0.01-0.15 in recall while satisfying the desired false alarm rate.

My second project, published at 2 workshops, addresses  both feature missingness and limited labels \citep{rath2022prediction, lobo2023semi}.  We proposed a solution based on the prediction constrained hidden Markov model (PC-HMM)\footnote{Code URL: \href{https://github.com/tufts-ml/pchmm-missing-data-limited-labels/}{github.com/tufts-ml/pchmm-missing-data-limited-labels/}}. This builds on previous work \citep{hughes2018semi, rabiner1986introduction, hopePredictionConstrainedHiddenMarkov2021} by adding a critical ability to handle missing data. Our proposed PC-HMM model \citep{rath2022prediction} is trained with an optimization objective that balances the HMM’s useful generative properties (handling missingness via marginalization) with the essential need for its learned representations to enable effective prediction in the clinical task. Experiments with our PC-HMM model showed absolute improvements of 0.02-0.08 in area under the precision-recall curve (AUPRC) over several state-of-the-art supervised and semi-supervised deep learning models at various levels of missing labels while predicting in-ICU mortality. Additionally, it is straightforward for domain experts to interpret our model due to its small number of parameters, each with a mechanistic role. Analysis revealed that our model enabled interpretability into patient groups that were at-risk. For example our model identified vulnerable patient groups as those with high blood-urea nitrogen (above 20 mg/dL) and high creatinine(above 1.45 mg/dL), which are common indicators of kidney problems \citep{baha2021effect}.

In the third project, we develop models for managing irregular time-series data, particularly when dealing with data-dependent missingness, a mechanism referred to as missing at random (MAR)\footnote{Code URL: \href{https://github.com/tufts-ml/data-driven-missingness-cru-irregular-timeseries}{github.com/tufts-ml/data-driven-missingness-cru-irregular-timeseries}}. This approach is critical for EHR time series, where  data collection is often based on data-driven clinical judgments regarding patient health changes. Many existing models either entirely neglect this assumption \citep{rubanova2019latent, bruun2020not}, or opt to instead input missingness indicators directly into deep learning networks \citep{che2018recurrent}. Such methods  are inadequate because they fail to learn the generative mechanism for missingness. While our earlier proposed PC-HMM model should be able to handle this scenario, we transition towards a continuous time model to accommodate data collected at irregular intervals. To handle irregular timing of measurements as well as data-dependent missingness scenarios, we extend a encoder-decoder stochastic differential equation (SDE) architecture named the continuous recurrent unit (CRU) \citep{schirmer2022modeling} to form a generative model that jointly produces an irregular time series of features and missingness indicators. In experiments across 3 health records datasets, our integrated model (CRU-FM) demonstrates lower extrapolation error than several alternative models, highlighting the importance of learning the data-dependent missingness mechanism in critical care time-series applications. 

 Ultimately, using the tools developed in this thesis, we envision a future where deployed ML models could use routine  physiological measurements (vitals, lab tests, and demographics) collected irregularly over time with missingness, to accurately identify patients at risk and generate alarms that clinicians could rely on to signal need for   early intervention, even when large labeled training sets are not easily available.

