\subsection{Potential for enhancing patient care with electronic health records and machine learning}
 Electronic health records (EHRs) fundamentally serve as clinical information systems designed to collect, maintain, and display electronic data that is gathered over time through the provision of healthcare services. In recent years, the expansive growth of EHRs and the development of sophisticated machine learning (ML) techniques have provided opportunities to enhance patient care in hospital settings, where the stakes are high and the demands are incessant. The integration of vast arrays of digital health information into EHR systems, ranging from patient vital signs and laboratory test results to medication orders and clinical notes, provides the ideal platform for deploying advanced ML algorithms. These algorithms aim to harness the predictive potential of complex, time-series data accumulated during patient hospital stays, with the promise of offering clinicians real-time insights that could anticipate patient trajectories, prevent adverse events, and tailor treatments more effectively. This capability to transform extensive data into actionable insights presents a vital bridge to the application of machine learning in creating robust early warning systems that significantly improve patient safety and treatment outcomes.

Predicting future patient outcomes from observed time series of labs and vitals in critical care settings has drawn significant recent research interest, catalyzed by the release of open-access datasets~\citep{johnson2020mimic, eICU}, benchmark classification tasks~\citep{purushothamBenchmarkingDeepLearning2018,harutyunyanMultitaskLearningBenchmarking2019a,wang_mimic-extract_2020,jarrett2021clairvoyance}, and compelling clinical deployments~\citep{sendakRealWorldIntegrationSepsis2020,hylandEarlyPredictionCirculatory2020}. Figure \ref{fig:ehr_ml_papers_by_year} shows the growing interest in this area of research. One particular application of interest to clinicians in developing such EHR based models is early warning alert systems. An early warning alert warns clinicians of impending deterioration in the patient's health. Several proxy labels are used to predict deterioration in the patient such as admission to the ICU, onset of a disease, increased length of stay at the hospital or in-hospital/in-ICU mortality. In our work, we focus on early prediction of mortality (as a binary outcome) to avoid focusing on a specific disease cohort. Later, in chapter \ref{chap:pchmm} we extend our models to predict the length of stay as an ordinal outcome, demonstrating their versatility across a range of clinical scenarios. 
 

\input{content/intro_figs/fig_ehr_ml_papers_by_year}

\subsection{Electronic health records datasets for predicting patient deterioration}
In this section, we elaborate on how we utilize 2 publicly available EHR datasets: MIMIC \citep{pollard2016mimic} and eICU \citep{pollard2018eicu}. These datasets are rich with time-stamped vital signs, laboratory measurements, medications, care plan information, diagnosis, patient history, microbiology cultures, and billing information.  In our work, we mainly focus on utilizing laboratory measurements and vital signs. 

\paragraph{Extracting time-stamped labs and vital measurements:} Deriving labs and vital measurements from the datasets is straightforward due to the standardized formats for data storage. For instance, MIMIC utilizes a system of internal identifiers, known as ITEMIDs, to categorize and store various types of clinical data including labs and vitals. Laboratory test results are recorded in the LABEVENTS table, which includes details such as the ITEMID—a unique identifier for each lab test—along with the $HADM\_ID$ (hospital admission ID), CHARTTIME (timestamp of the lab result), and the lab values themselves. The VALUENUM field captures the numerical results of the test, and VALUEUOM denotes the unit of measurement. Similarly, vital signs are meticulously logged in the CHARTEVENTS table, a comprehensive repository that chronicles all charted data, including routine vital signs like heart rate, blood pressure, respiratory rate, temperature, and oxygen saturation. Each entry in this table is tagged with an ITEMID, indicating the type of data recorded, complete with timestamps, values, and occasionally additional information such as whether the data was manually entered or automatically recorded.

\paragraph{Deriving outcomes:} Outcomes such as mortality and length of stay in the hospital can often be directly obtained from the ADMISSIONS table, which contains entries for the date and time of death during the hospital admission. Features and outcomes for each patients are linked from multiple tables (such as PATIENTS, ADMISSIONS, and ICUSTAYS) to get a comprehensive view of the patient’s timeline at the hospital.


\subsection{Complications in operationalizing EHR based ML models in clinical settings}
Despite the public availability of large scale EHR datasets, the practical application of adopting ML models in clinical settings is often hampered by significant hurdles. First among these is the prevalence of missing data in EHRs—an inevitable consequence of the sporadic nature of healthcare monitoring and data entry practices. Vital signs or lab results might be recorded irregularly, entries may be omitted during acute patient care, or data may simply be lost or incorrectly entered. Roughly $60-80\%$ of hourly measurements are missing in datasets such as MIMIC and eICU \citep{purushotham2018benchmarking, rath2022prediction}. This missing data poses a critical challenge for ML models, which typically require complete datasets to function optimally. Without robust mechanisms to handle gaps in data, the accuracy of predictions and the effectiveness of decision-support systems can be severely compromised.

Moreover, the labels required to train these predictive models—such as indications of disease onset or patient outcomes—are often sparse. Label sparsity occurs because the intensive labor required to annotate medical records accurately is both time-consuming and costly, restricting the availability of well-labeled datasets \citep{lam2021semisupervised, wessler2023automated}. Given the constraints on label sparsity, there is a compelling need for machine learning models that can learn effectively from limited labeled data, making semi-supervised learning approaches particularly valuable in the healthcare context. Semi-supervised learning models are designed to capitalize on the large amounts of unlabeled data that are typically available in EHR systems, using them to enhance learning and improve model accuracy even when comprehensive, expert-labeled datasets are not available.

Another significant challenge is the high rate of false alarms generated by many ML systems. These false alarms can arise from over-sensitive models that err on the side of caution, flagging potential issues that do not pose real threats to patient health. While initially appearing as a cautious strategy, a high rate of false alarms can lead to 'alarm fatigue' among healthcare providers, reducing their attention to alerts and decreasing their trust in the system's accuracy. This can ultimately undermine the utility of ML applications in clinical settings, as healthcare providers begin to disregard alerts, potentially missing critical warnings that require immediate attention \citep{sendelbach2013alarm}.

Collectively, these challenges—missing data, sparse labels, and a high rate of false alarms—degrade the performance and trustworthiness of ML-based predictive systems in healthcare. Addressing these issues is not merely a technical challenge but a prerequisite for ensuring that the promise of ML to enhance patient care can be fully realized in a practical, sustainable, and clinically beneficial manner. As such, developing robust ML models that can effectively navigate these complexities is critical for the advancement of intelligent healthcare systems that clinicians can rely on to support their decision-making in real-time.

\subsection{Outline} Having provided a background on clinical prediction models and obstacles in operationalizing these models in practice, below is an outline on the subsequent chapters in this thesis. 

Chapter 3 delves into optimizing early warning systems to control false alarms through a minimum precision constraint. It elaborates on new tractable bounds for maximizing recall and empirical success in realistic clinical tasks. Chapter 4 introduces methods for predicting patient outcomes from time-series EHR with missing entries and limited labels by employing a semi-supervised Hidden Markov Model to enhance prediction accuracy and manage data sparsity. Chapter 5 discusses probabilistic modeling of data-driven missingness in irregular time series medical records, proposing a new model that better handles the irregularity and missingness common in electronic health records. Finally, Chapter 6 offers a discussion and outlook, reflecting on the impacts and future directions of the research. This structure underscores a progressive deepening into specific challenges and innovations in machine learning for healthcare monitoring.