
\textbf{Loose bound justifies per-site mean ranking.} A natural first guess for ranking is the per-site mean: $\bm{\bar{r}} = \mathbb{E}_{p_{\phi}}[\bm{y}]$. We can show this is justifiable way to minimize expected loss on a simplistic upper bound on BPR.
Assume there exists an \emph{upper limit} $U$ such that for all $s$, we can guarantee $U \geq y_s$.
%For example, we may know that per-site counts will be smaller than overall population size. 
The sum over any $K$ entries of vector $\bm{y}$ in the denominator of BPR is then bounded by $K \cdot U$. Plugging this bound into the minimize expected loss problem and simplifying with linearity of expectations yields
% \begin{align}
% \label{eq:bound_on_bpr}
%     \text{BPR}(\bm{\theta}, \bm{y}) 
%     \geq \frac{
%     \bm{y} \circ \textsc{TopKMask}( \theta, K)
%     }{
%         \sum_{s=1}^S y_s
%     }
%     \geq \frac{
%     \bm{y} \circ \textsc{TopKMask}( \theta, K)
%     }{
%         S \cdot U
%     }, \qquad \text{when}~ U \geq y_s \forall s
% \end{align}
% Optimizing our how-to-act problem using this bound yields
\begin{align}
    \bm{r}^* \gets \argmin_{\bm{r}} -
    \underbrace{
    \frac{
    \mathbb{E}_{p_{\phi}} [ \bm{y} ]
    }{
        K \cdot U
    } \cdot \textsc{TopKMask}( \bm{r}, K)
    }_{\leq \text{BPR}(\bm{r}, \bm{y})}
\end{align}
Many solutions exist: any vector $\bm{r}^*$
that satisfies
\[
    \textsc{TopKMask}(\bm{r}^*, K) = \textsc{TopKMask}(\mathbb{E}_{p_{\phi}}[\bm{y}], K)
\]
can be an argmin. One valid solution here is the per-site mean $\bm{\bar{r}}(\phi) = \mathbb{E}_{p_{\phi}}[\bm{y}]$. However, this solution is optimal for a potentially quite loose bound on BPR that approximates the denominator with the constant $K \cdot U$.

\textbf{Tighter bound suggests the ratio estimator for ranking.} 
Instead of bounding with constant $K \cdot U$, we can upper bound the denominator in Eq.~\eqref{eq:loss_bpr_how_to_rank} by summing over all $S$ terms instead of the top $K$:
$\sum_{s \in \textsc{TopKIDs}(\bm{y})} y_s \leq \sum_{s=1}^S y_s = \bm{1} \cdot \bm{y}$. This bound is tighter when the sum of all entries is less than $K \cdot U$, which is typically true of sparse $\bm{y}$ vectors in our applications. Our how-to-rank problem then becomes
\begin{align}
    \argmin_{\bm{r}} - \underbrace{
    \mathbb{E}_{\bm{y} \sim p_{\phi}} \left[
    \frac{
    \bm{y} 
    }{
        \bm{1} \cdot \bm{y}
    }
    \right] \circ \textsc{TopKMask}( \bm{r}, K)
    }_{\leq \text{BPR}(\bm{r}, \bm{y})}
    % \\
    % %  &= \sum_{s=1}^S 
    % %  - \mathbb{E}_{y \sim p_{\phi}} \left[
    % % \frac{ y_s }{\sum_{u=1}^S y_u}
    % % \right]
    % \cdot \textsc{TopKMask}( \theta, K)_s
\end{align}
Again, we solve via any vector $\bm{r}^*$ whose top-K binary vector
$ \textsc{TopKMask}(\bm{r}^*, K)$ equals $\textsc{TopKMask}( \mathbb{E}_{p_{\phi}} [
    \frac{ \bm{y} }{\bm{1} \cdot \bm{y}}
    ], K )$.    
One valid solution is the expected ratio of vector $\bm{y}$ to its sum, which we nickname the \emph{ratio} estimator
\begin{align}
\label{eq:ratio_estimator_for_ranking}
\bm{r}^*(\phi) = \mathbb{E}_{\bm{y} \sim p_{\phi}}\left[ \frac{\bm{y}}{\bm{1} \cdot \bm{y}} \right]
\approx \frac{1}{M} \sum_{m=1}^M \frac{ \bm{y}^{(m)}}{ \bm{1}\cdot\bm{y}^{(m)} }.
\end{align}
This ranking is \emph{distinct} from the per-site mean: there exist fixed models $\phi$ where the ratio estimator and the per-site mean would select different subsets of the same $S$ sites.
A natural question is, why not use a ranking estimator which more closely resembles BPR, such as $\bm{r} = \mathbb{E}[ \frac{ \bm{y} }{ \text{TopKMask}(\bm{y},K)} \cdot \bm{y}  ]$. 
In practice, this estimator is equivalent to the ratio estimator in expectation, and will select the same top-K locations. Accordingly, we use the ratio estimator for its improved computational speed and simplicity.
See App.~\ref{supp:ranking_demo} for concrete cases where the ratio estimator earns BPR 2.5x to 5x higher than the per-site mean, even when all estimators know the true data-generating model.

When exact computation of this expectation is not easy, we recommend a Monte Carlo approximation using $M$ samples $\{ \bm{y}^{(m)} \}_{m=1}^M$ drawn iid from $p_{\phi}$, as in Eq.~\eqref{eq:ratio_estimator_for_ranking}. 
This is a stochastic estimator; rankings can differ across repeat trials if $M$ is not large enough.
