To gain insight, we designed a toy classification task with 2 features where the solutions favored by different objectives can be easily visualized.
Our goal is to train a logistic regression (LR) classifier whose decisions maximize recall subject to a minimum precision constraint of $\alpha = 0.9$.
We illustrate the advantage of our logistic sigmoid bound empirically by comparing it with 5 other objectives.
Fig.~\ref{fig:toy_data_precision_90} shows all training examples ($N_+ = 120, N_- = 450$) together with the learned linear decision boundaries for four possible optimization strategies: BCE with post-hoc threshold selection, \citet{ebanScalableLearningNonDecomposable2017}'s hinge-bound approach, \citet{fathonyAPperfIncorporatingGeneric2020}'s adversarial prediction (AP) method, and our proposed sigmoid bound approach. We further compare 2 other methods, class-weighted cross entropy and a log loss surrogate bound in Sec.\ref{sec:other_possible_bounds}.

% Even with post-hoc search, BCE cannot achieve the desired precision of 0.9.
Although the BCE solution might look reasonable, recall that our specific goal is to maximize recall subject to a precision of at least 0.9. The BCE solution gets only 0.69 precision even with post-hoc threshold search, and thus remains quite far from the desired precision of 0.9.
Similarly, the hinge bound achieves only 0.79 precision, also falling short of our goal.
Both the AP method and our approach can satisfy the desired precision constraint. 
However, our proposed sigmoid bound achieves noticeably better recall of 0.24, which is over double the 0.11 recall found by AP.

To investigate bound quality, we verified that even after initializing at the sigmoid bound's better solution, optimizing Eq.~\eqref{eq:standardized_objective_UNconstrained} using the hinge bound prefers to wander away, suggesting the inadequacy of this loose bound (see our code repository for a reproducible example).

Finally, we performed an exhaustive grid search of the 3 LR parameters (two weights, one bias), optimizing the ideal objective in Eq.~\eqref{eq:max_recall_st_precision}. (This search is only feasible in low-dimensional problems.) The grid search optimum (not shown) reached precision 0.9 and recall 0.24, suggesting our sigmoid bound's precision 0.96 and recall 0.20 is close to the optimal to beat for this dataset.
